A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-10-16 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we will...

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions to be novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
are essentially composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
reduction operations.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures", and we'd be happy to change the 
title should our submission be accepted.

### 2012-10-17 doc/asplos2013/response.txt

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
are essentially composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
reduction operations.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be willing to change our
title to clarify our meaning if the submission is accepted.

Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node) unlike our 
events which are valid in every address space on distributed memory machines.  Furthermore, the
synchronization primitives in OpenCL do not use events and cannot compose with other operations.

Reviewer 3 expressed doubts about the performance of the Legion version of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which has been accepted and will appear at Supercomputing 2012) 
reports speedups over hand-coded versions of the applications.  In this paper, we aimed to quantify 
exactly what fraction of the speedup from [4] results from the composable, asynchronous nature of our 
runtime system.  To eliminate any performance benefits achieved by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
written in Legion instead of hand-coded implementations.  The Fluid and AMR applications had hand-coded 
reference implementations written by others.  In both cases, the architects of those programs chose to 
express their computation in a bulk-synchronous model indicating that the programs are a good fit
for bulk-synchronous execution.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

### 2012-10-17 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we will...

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the deferred application of these
reduction operations.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
title to clarify our meaning if the submission is accepted.

@@

Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node) unlike our 
events which are valid in every address space on distributed memory machines.  Furthermore, the
synchronization primitives in OpenCL do not use events and cannot compose with other operations.

Reviewer 3 expressed doubts about the performance of the Legion version of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which has been accepted and will appear at Supercomputing 2012) 
reports speedups over hand-coded versions of the applications.  In this paper, we aimed to quantify 
exactly what fraction of the speedup from [4] results from the composable, asynchronous nature of our 
runtime system.  To eliminate any performance benefits achieved by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
written in Legion instead of hand-coded implementations.  The Fluid and AMR applications had hand-coded 
reference implementations written by others.  In both cases, the architects of those programs chose to 
express their computation in a bulk-synchronous model indicating that the programs are a good fit
for bulk-synchronous execution.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 described deferred locks as having the same semantics as I-structures from the dataflow
literature.  This is incorrect.  I-structures support a producer-consumer semantics which is isomorphic
to the functionality provided by events in our runtime (we do not claim the semantics of events to
be novel).  Unlike the producer-consumer semantics supported by I-structures, deferred locks support
unordered atomic semantics, which allows programmers to describe properties like serializability
without the requirement of ordering.  We are unaware of any feature of dataflow implementations
that support a similar semantics to deferred locks.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads (except on GPUs where 
threads are lightweight).  Instead our system provides lighter-weight tasks which the programmer
can launch at a much finer granularity with incurring the overhead of a thread.  Due to the minimal
overhead of tasks, programmers can express a much finer granularity of parallelism which the runtime
can then leverage if there are available hardware resources.  Existing programming systems like MPI
and P-threads entail too much overhead to leverage this granularity of task-parallelism.

Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experiencing writing programs to our runtime system, we've found that
the number of expected arrivals at a barrier is not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
being able to dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  GPUs are just another kind
of processor in our system and GPU memories (global, zero-copy, etc) are just another kind of memory with
regards to the runtime.  Copies of data between GPU and CPU memories are no different than copies between
any other pair of memories and must be explicitly made by the programmer.  Tasks launched on a GPU processor
invoked a kernel on the GPU but otherwise have the same semantics as CPU task.  One restriction is 
runtime calls are not currently available inside of tasks running on the GPU.  We left this aspect of 
the implementation for future work and owing to this asymmetry opted not claim heterogeneity as one
of our primary contributions.

### 2012-10-17 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we will...

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most also satisfy our criterion of foldability
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

@@

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
title to clarify our meaning if the submission is accepted.

Reviewer 2 noted that OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events are only valid in a single address space (e.g. node),
whereas our events which are valid in every address space on distributed memory machines.  Furthermore, the
synchronization primitives in OpenCL do not use events and cannot compose with other operations.

Reviewer 3 expressed doubts about the performance of the Legion version of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
written in Legion instead of the reference implementations.

@@

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 was curious about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics which is much closer
to the functionality provided by events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow implementations that support a similar semantics to deferred locks.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
Instead our system provides lighter-weight tasks which the programmer
can launch at a much finer granularity with incurring the overhead of a thread.  Due to the minimal
overhead of tasks, programmers can express a much finer granularity of parallelism which the runtime
can then leverage if there are available hardware resources.  Existing programming systems like MPI
and P-threads entail too much overhead to leverage this granularity of task-parallelism.

@@

Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experiencing writing programs to our runtime system, we've found that
the number of expected arrivals at a barrier is not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
being able to dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and launches
asynchronous CUDA grids for each task.  GPU memories (framebuffer and zero-copy) are exposed in the same way
as CPU main memory and the globally accessible "GASNet" memory.  Copies of data between GPU and other
memories are explicitly requested by the programmer (like all other copies), and are performed with
asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, which are exposed
in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we address
the comments and questions most frequently raised by the reviewers.

@@

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
title to clarify our meaning.

Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events which are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and cannot compose with other operations the way the synchronization mechanisms in our runtime can.

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
written in Legion instead of the reference implementations.

@@

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 was curious about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics which is much closer
to the functionality provided by events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow implementations that support a similar semantics to deferred locks.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
Instead our system provides lighter-weight tasks which the programmer
can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
and P-threads entail too much overhead to leverage this finer-grained task parallelism.

@@

Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experiencing writing programs to our runtime system, we've found that
the number of expected arrivals at a barrier is not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
being able to dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
launched on a GPU processor results in an asyncrhonous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
data between GPU and other memories are explicitly requested by the programmer (like all other copies), 
and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
which are exposed in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was the use of "commutativity".  The traditional use of these terms is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be willing to change the 
title to clarify our meaning.

Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events which are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.

@@

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous version of the applications from [4] 
written in Legion instead of the reference implementations.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal work to dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics which is much closer
to the functionality provided by events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

@@

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
Instead our system provides lighter-weight tasks which the programmer
can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
and P-threads entail too much overhead to leverage this finer-grained task parallelism.

Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experiencing writing programs to our runtime system, we've found that
the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the feature of
being able to dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

### 2012-10-23 doc/types-pldi2013/langdef.tex
operational semantics rewrite

Core Legion's operational semantics rules have the following form:
$$\opsenv{e}{v}{E}$$
where: 
\begin{itemize}
\item $M$ is a mapping from logical regions $r_i$ to {\em physical regions} $\rho_i$, which are sets of concrete memory locations.
We write $\hat x = M \llbracket x \rrbracket$ for the application of $M$ to other objects (e.g., types), 
replacing all free logical region names $r_i$ in $x$ with corresponding physical regions $\rho_i$.  
(Note $M \llbracket x \rrbracket$ is ill-formed if $x$ contains a free logical region name $r_j \not\in dom(M)$.)

\item $L$ is a mapping of local variables to (immutable) values.

\item $H$ is a {\em heap typing}, assigning an invariant type to every memory location.  The heap typing
serves two important roles.  First, it is used to bound the values that can be used for reads when 
concurrent operation may make the exact value uncertain.  Second, it greatly simplifies the consistency
check of pointer values, allowing a simple test against the heap typing for a location in lieu of a
potentially unbounded walk through the store for complicated pointer data structures.

\item $S$ is a store, mapping memory locations $l_i$ to values.

\item $C$ is a {\em clobber set}, memory locations that may be accessed by concurrently executing operations.

\item $e$ is an expression.

\item $v$ is the result of evaluating $e$.

\item $E$ is a {\em memory trace} of the execution, a list of {\em memory operations}
in the order performed.  A memory operation $\epsilon$ is either
$read(l, c, v)$ (a read of location $l$ with coherence mode $c$, returning value $v$),
$write(l, c, v)$ (a write to location $l$ with coherence mode $c$, storing value $v$),
or $reduce_{id}(l, c, v)$ (a reduction into location $l$ with coherence mode $c$ using an
associative function $id$ with value $v$).  Coherence modes are one of ({\em excl}, {\em atomic}, {\em simult}).
\end{itemize}

Helper functions are used in some rules:
\begin{itemize}
\item $apply(S, E)$ computes the store that results from applying memory trace $E$ to initial store $S$.

\item $mark\_coh(E, \hat Q)$ records the coherence of operations in trace $E$ according to coherence annotations $\hat Q$.

\item $valid\_interleave(S, C, E', \oton{E})$ is true if $E'$ is a valid interleaving
of the memory traces $\oton{E}$.
\end{itemize}
Figure~\ref{fig:opsemfns} defines {\em apply} and {\em mark\_coherence}.  The precise definition of
{\em valid\_interleave} is given in Section~\ref{sec:coherence}.  For now, we need only that a valid interleaving
is at least an interleaving, containing each constituent trace as a subsequence and no other operations.

Some rules modeling concurrent execution are locally
non-deterministic, guessing values that result from
concurrent tasks.  For example, the rule for {\tt read(e)} returns an
arbitrary value $v'$ of the correct type if the location being read is
in the clobber set $C$, the set of locations concurrent tasks may
clobber (i.e., update).  The guess is eventually checked for
consistency by the $valid\_interleave$ predicate of some enclosing
expression, at a point where the task that wrote $v'$ is also local to
the expression being evaluated.

Another prescient operation is {\tt newcolor r}, which guesses initial
memory locations and their colors for a partition.  These locations
support partition-then-allocate, ensuring the memory mapping $M$ has
locations for the new subregions that can be allocated by {\tt new}.
As is usual, memory allocation is also non-deterministic, choosing any
location of the correct type in the region.  For simplicity we do not
require locations to be ``fresh''; our soundness results hold even if
{\tt new T@r} returns locations already in use, provided the heap
typing of the location is correct.

The most involved rule in Core Legion's semantics is for function calls.
The only regions accessible to the callee (through $M'$) are the region arguments.
Also, if any regions are used with relaxed coherence (either {\em atomic} or {\em simult} instead of
{\em excl}), those locations are added to the clobber set, allowing the
impact of concurrent modifications of the store to be handled by the caller.  
%This is done by
%annotating the called function's memory trace with the relaxed coherence on a per-location basis
%and using the {\em valid\_interleave} predicate to control what parallel execution is allowed.

>>>>

Core Legion's operational semantics rules have the following form:
$$\opsenv{e}{v}{E}$$
and specify that, given a certain environment, the evaluation of expression $e$ yields a value $v$ and a
{\em dynamic memory trace} $E$ which is a sequence of all heap operations (reads, writes, reductions) performed
during the evaluation, including their location and the value read or written.

In addition to the usual value mappings for local variables $L$ and store (i.e. heap) locations $S$, the Legion
operational semantics environments includes an immutable heap typing $H$ that maps locations to types, and a region
mapping $M$ that maps logical regions $r_i$ to physical regions $\rho_i$, which are sets of concrete memory
locations.

The final element of the environtment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithms
used by the Legion runtime satisfy these constraints, guaranteeing the soundness of the Legion programming model.

The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.

The first use of the region mapping $M$ occurs in the {\em new} and {\em downregion} expressions.  The new
expression is constrained to select a location (with the right heap typing) from the set that the specified logical
region has been mapped to.  (Although one would expect the new expression to also return a location that is not
currently in use, our soundness result does not require this additional constraint, so we have omitted it for
simplicity.)  Similarly, the downregion expression checks whether a location is within the set of locations the
logical region(s) have been mapped to, returning {\em null} if not.  As discussed above, the corresponding 
{\em upregion} expression checks are performed statically.

The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
existing coloring and returns one in which the specified location is given the specified color, replacing an
existing coloring for that location (if present).  The behavior of the newcolor expression is a bit trickier.
Since Legion supports the {\em partition-then-allocate} model in which the {\em new} expression may be called
on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
guaranteeing that any explicit colorings performed by the program are honored.

The semantics for the {\em pack} and {\em unpack} expressions are straightforward.  Packing a region relationship
is just a matter of capturing the physical regions to which the logical regions are currently mapped, whereas 
unpacking takes those physical regions and gives them fresh logical names within the evaluation of the body.

Finally, although the semantics for calls to subtasks appear complicated at first glance, the bulk of the rules
are just handling the evaluation of each actual argument expression and the creation of the called subtask's 
environment (with potentially different names for logical regions).  The key difference in Legion task calls is the
handling of relaxed coherence.  Any location that falls within
a logical region whose coherence has been relaxed (via {\em atomic} or {\em simult}) is added to the clobber set
$C'$, allowing non-deterministic evaluation within the subtask with respect to those locations.  All actual accesses
to those locations within the resulting memory trace ($E_{n+1}$) are marked with the corresponding coherence mode,
allowing the {\em valid\_interleave} check within the caller's environment to verify the consistency of the overall
memory trace.
